DaLi: Database as a Library

### 2017-02-18 paper/semantics.tex
Unique LCA theorem added.

\begin{definition} [\bfseries External Locus]
Given a branch $b$ and a version $v\in b$, a version $v_o \not\in b$
is an external locus of $v$ iff it is a least ancestor of $v_o$
outside of $b$. That is, $\under{H}{v_o \preceq v}$, and there does
not exist a $v_o' \not\in b$ such that $\under{H}{v_o' \preceq v}$ and
$\under{H}{v_o \preceq v_o'}$. We lift the definition to the level of
branches; $v_o$ is an external locus of a branch $b$ iff it is an
external locus of its latest version.
\end{definition}

Note that all branches (except $H(t_{\top})$) begin with a \C{FORK}
tag, hence every version has at least one external locus. We later
prove that in the histories generated by the rules in
Fig.~\ref{fig:opsem}, each version (hence, each branch) has exactly
one external locus.

@@

In \rulelabel{E-Run} case, initial version is the external locus of
the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
which creates the branch $H(t')$ by forking from $H(t)$, the latest
version on $H(t)$ is the external locus of the first version on
$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
external locus as the previous latest version, which, premise says, is
unique. No new version is created in \rulelabel{E-Pull}. In
\rulelabel{E-Pull-Wait}, new version is created by merging the latest
version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
Premise says that $v$ has a unique external locus (call it $v_o$). The
new version $v_m$ inherits $v_o$ as an external locus, and also has
$v'$ as another external locus. However, since $\under{H}{v' \mbleto
v}$, we have $\under{H}{v_o \preceq v'}$. Hence $v'$ is the only
external locus.
\qed
\end{proof}

@@

The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
Let $v_1$ be the previous version on the branch, and let $v_2$ be any
other version. The premise is that $v_1$ and $v_2$ have a unique LCA
$v_0$. Observe that if $v_0 \neq v_1$, then LCA of $v$ and $v_2$ must
also be an LCA of $v_1$ and $v_2$, hence LCA of $v$ and $v_2$ is
unique. If $v_0 = v_1$, then LCA of $v$ and $v_2$ is $v_1$, hence
unique. Thus, uniqueness of LCA is preserved. The reasoning for
\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.

The \rulelabel{E-Pull-Wait} case is where the merge happens. 
\end{proof}

>>>>

\begin{definition} [\bfseries Internal and External Ancestors]
Given a branch $b$ and a version $v\in b$, an internal ancestor of $v$
is an ancestor from the same branch $b$. An external ancestor of $v$
is an ancestor from a different branch $b'\neq b$. 
\end{definition}

\begin{definition} [\bfseries Least Internal and External Ancestors]
Given a branch $b$ and a version $v\in b$, least internal ancestor is
the immediate ancestor version ($v_i$) of $v$ on $b$. Least external
ancestor ($v_o$) is an external ancestor of $v$ that is not an
ancestor of any other external ancestor of $v$. That is,
$\under{H}{v_o \preceq v}$, and there does not exist a $v_o' \not\in
b$ such that $\under{H}{v_o' \preceq v}$ and $\under{H}{v_o \preceq
v_o'}$. 
% We lift the definition to the level of branches; $v_o$ is an
% external locus of a branch $b$ iff it is an external locus of its
% latest version.
\end{definition}

Note that all branches (except $H(t_{\top})$) begin with a \C{FORK}
tag, hence every version has at least one least external ancestor. We
later prove that in the histories generated by the rules in
Fig.~\ref{fig:opsem}, each version has exactly one least external
ancestor. We call the unique least external ancestor of a version as
its \emph{external locus}. Least internal ancestor, by definition, is
unique, which we sometimes call \emph{internal locus}.

@@

In \rulelabel{E-Run} case, initial version is the unique external locus of
the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
which creates the branch $H(t')$ by forking from $H(t)$, the latest
version on $H(t)$ is the external locus of the first version on
$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
external locus as the previous latest version, which, premise says, is
unique. No new version is created in \rulelabel{E-Pull}. In
\rulelabel{E-Pull-Wait}, new version is created by merging the latest
version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
Premise says that $v$ has a unique external locus (call it $v_o$). The
new version $v_m$ inherits $v_o$ as a least external ancestor, and
also has $v'$ as another least external ancestor. However, since
$\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
have $v'$ as the the external locus.
\qed
\end{proof}

%% If we are allowed to merge earlier versions on branches, then we
%% have a possibility of utmost two LCAs.

@@

The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
Let $v_1$ be the previous version on the branch, and let $v_2$ be any
other version. The premise is that $v_1$ and $v_2$ have a unique LCA
$v_0$. Observe that $v$'s internal locus is $v_1$ and its external
locus is same as that of $v_1$. Hence, all ancestors of $v_1$
(including itself) are also ancestors of $v$. Moreover, $v$ has no new
ancestors except itself, which is the latest version being added.
Hence, any common ancestor of $v_1$ and $v_2$ is also a common
ancestor of $v$ and $v_2$, while there cannot be new common ancestors.
Hence LCA of $v$ and $v_2$ is unique. The reasoning for
\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.

The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
$v_m$ is merging into version $v$ to produce the next version $v_m$.
Let $v_2$ be any version. From premise, we know that $v$ and $v_2$
have a unique LCA (call it $v_0$), and $v'$ and $v_2$ have a unique
LCA (call it $v_1$). But, we know that $v_2$ has a unique external
locus, which leads to the following cases:
\begin{itemize}
  \item Both $v_0$ and $v_1$ are external ancestors of $v_2$. Hence,
  $\under{H}{v_0 \preceq v_1} \disj \under{H}{v_1 \preceq v_0}$.
  Since\under{H}{v_0 \preceq v_m} \disj \under{H}{v_1 \preceq v_m}, 
  $v_2$ and $v_m$ have a unique LCA.

  \item $v_0$ is an internal ancestor of $v_2$, hence an external
  ancestor of $v$, hence an external ancestor of $v_m$. $v_1$ is
  an ancestor of $v'$, hence also an external ancestor of $v_m$. 
\end{itemize}

As proved
before, the merged version ($v_m$) has a unique external locus, which
is the merging version ($v'$).
\end{proof}
